Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2019-02-28 manuscript/homophily_manuscript.Rmd
Ready for second resubmission

However, we believe that earlier studies of gender homophily were hindered by a largely unacknowledged statistical issue that we will refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Whalund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues even if no gender preferences exist, or if the true preference is for opposite-gendered colleagues ('gender heterophily'). For example, a sample of literature containing a mixture of bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers usually collaborate within their own discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplementary Material). Papers with 3, 4 or $\ge5$ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3 author) and larger (5+ authors) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true. Essentially, this issue arises because a person cannot be their own co-author. In a small dataset comprising $m$ men and $f$ women, a man can co-author with $m - 1$ men while women can co-author with $m$ men. Thus, the null expectation for $alpha$ is a negative number -- potentially a large one if $m$ and $f$ are very small.

@@

# Acknowledgements
CM was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions). We thank Devi Stuart-Fox and Dominique Potvin for helpful discussion.

>>>>

However, we believe that earlier studies of gender homophily were hindered by a largely unacknowledged statistical issue that we will refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Wahlund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues even if no gender preferences exist, or if the true preference is for opposite-gendered colleagues ('gender heterophily'). For example, a sample of literature containing a mixture of bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers usually collaborate within their own discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplementary Material). Papers with 3, 4 or $\ge5$ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-author papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3-author) and larger (5+ author) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true. Essentially, this issue arises because a person cannot be their own co-author. In a small dataset comprising $m$ men and $f$ women, a man can co-author with $m - 1$ men while women can co-author with $m$ men. Thus, the null expectation for $\alpha$ is a negative number -- potentially a large one if $m$ and $f$ are very small.

@@

# Acknowledgements
We thank Devi Stuart-Fox and Dominique Potvin for helpful discussion.
